Chasing \rainbows{} with the Habitable Worlds Observatory

### 2023-04-19 src/tex/ms.tex
[showyourwork] overleaf sync

\usepackage{siunitx}
\sisetup{
    range-units = single,
    range-phrase = {--},
    separate-uncertainty = true,
    multi-part-units = brackets,
    product-units = brackets,
    detect-weight = true,
    per-mode = symbol,
    mode = text,
}
\DeclareSIUnit\parsec{pc}
\DeclareSIUnit\au{au}
\DeclareSIUnit\mas{mas}

@@

% Define macros
\newcommand{\IWA}{\ensuremath{\mathrm{IWA}}}
\newcommand{\hwo}{HabWorlds}

@@

\begin{abstract}
NASA recently announced the Habitable Worlds Observatory (\hwo), a coronagraphic mission to detect rocky planets in habitable zones, establish their habitability, and search for biosignatures. 
%Numerous instrumental and observational challenges have still to be overcome to design a coronagraphic mission with that goal. After a research phase aiming at the detection of a few tens of exo-Earth in the habitable zone, the Habitable Worlds Observatory will use its spectroscopic and polarimetric capabilities to probe these planets' atmosphere and surface properties. 
Surface liquid water is central to the definition of planetary habitability.
%
Photometric and polarimetric phase variations can show the presence of specularly reflecting oceans on exoplanets, in addition to being sensitive to rainbows and other cloud scattering effects. 
%
Direct imaging missions are optimised to detect planets near quadrature, which may obscure some of the phases at which these features are strongest. 
%
The range of scattering phases accessible for an exoplanet can be limited by its orbital inclination or the coronagraph's inner working angle. 
%
Planets that orbit close to edge-on have accessible phase angles limited by obscuration by the coronagraph. 
%
We use the list of target stars for the Habitable Worlds Observatory to estimate the number of exo-Earths that could be searched for non-Lambertian scattering phenomena. 
%
We find that exo-Earths in systems listed in this Habitable Worlds Observatory catalog will have a detectable Rayleigh scattering peak. 
%
The glint signature at planetary phase angles of $\sim$50--70$^\circ$ 
%\sim$110--130$^\circ$ 
would be accessible to the Habitable Worlds Observatory in $\sim$14 systems, and the rainbow of water clouds at phases of 140--160$^\circ$ would be accessible in $\sim$42 systems, assuming a 62\,mas inner working angle (\IWA{}; 3$\lambda/D$ for a 6-m telescope at 600nm).
%
Increasing the \IWA{} to 2$\lambda/D$, increases the number of systems with detectable rainbows and glints by factors of roughly 2 and 3, respectively.
%
The number of systems in which an exo-Earth could be searched for these scattering phenomena increases with orbital eccentricity and scales approximately inversely with \IWA{}.    \end{abstract}

@@

%--------------------------------------------------------------------------------
\begin{figure*}%[t]
   \centering
   \includegraphics[width=\linewidth]{figures/Fig1IWAPol.png}
   \caption{An illustration of phase curves for two homogeneous planets in an edge on system with oceans and an Earth-like atmosphere. The cloudy scenario includes an Earth-like water cloud deck at low altitudes. Below the plot, the features that the phase angles and scattering angles are most sensitive to are shown. In reality, as can be seen by the examples above, these features overlap.  Throughout the paper we approximate feature sensitivity ranges as those that capture the rise an fall of peaks in polarized light for most habitable zone species in the UV, Optical, and IR (scattering angles: glint $\sim$50--70$^\circ$; rainbow $\sim$140--160$^\circ$}; Rayleigh narrow of quadrature, and other forward and backscattering effects at the widest and narrowest scattering angles).
    \label{fig:bottplot}
\end{figure*}
%--------------------------------------------------------------------------------

Signatures of habitability, from ocean glint to optical effects from condensates --- like the rainbow--- should be some of the most distinct signals in an exoplanet's reflected light phase curve, particularly in polarization.
%
However, depending on system architectures, some of these features can be obscured by the inner working angle of a coronagraph.

Liquid surface water is closely related to planetary habitability because it is essential for life as we know it.
%
To first order, the presence of liquid surface water can be predicted by the flux a planet receives from its star.
%
The habitable zone (HZ) is defined as the range of semi-major axes around a given main sequence star for which a rocky planet (or moon) may have the right temperature to maintain liquid surface water \citep{kasting93}. 
%
However, there are many factors besides the stellar flux that influence a planet's surface temperature and ability to hold liquid surface water, such as the atmospheric surface pressure and e.g.\ geological activity. 
%
Establishing the presence of liquid surface water on an exoplanet would therefore provide important context for the interpretation of possible biosignatures.
%
Providing additional context to the habitability of an exoplanet is the presence of a water-cycle as evidenced by the identification of a rainbow and other optical effects that can be attributed to water clouds.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Exoplanet Phase Curves}

@@

\subsection{Ocean Glint}

%As ocean glint is one of the few sources of specular reflection and is strongly polarized, it provides one of the best means to detect a signature of habitability in exoplanet phase curves. 
Ocean glint is the specular (Fresnel) reflection of light off smooth liquid surfaces.
%
In the total flux the glint signal will be most prominent at small planetary phase (large scattering) angles.
%
The maximum polarization occurs at the Brewster angle determined by the refractive indices at the interface; between air and liquid water this occurs at 53$^\circ$ \citep[a scattering angle of 127 degrees; see e.g. ][]{2008Icar..195..927W}.
%
This angle therefore varies slightly with wavelength and with sea and atmosphere composition.
%
In general, the glint on a planet will be strongest at red wavelengths where the Rayleigh scattering in the planet's atmosphere and attenuation through the atmosphere are smallest \citep{Zugger_2011} and, conveniently, where the latitude-albedo effect (a degeneracy between polar brightness and glint) is also small \citep{2012ApJ...752L...3C}.
%
The strengthened glint signature at longer wavelengths has indeed been observed in Earthshine measurements \citep{Emde2017,sterzik2019, takahashi2021}.
%
{\color{red}{check the next sentence:}}
The broadening of the glint feature in the presence of wind-driven waves is well-described by the Cox-Munk solution \citep{CoxMunk1954}, shifts in both polarized and unpolarized light, and broadens across scattering angles in polarized light \citep{kopparla2018, Zugger_2010, treesandstam2019, trees2022}.

The glint provides evidence of liquid surface bodies, which taken in the context of the temperature and types of condensates, may be interpreted as evidence for the presence of liquid water.
%
The forward scattering of light by cloud particles could be confused with glint \citep{Robinson_2010}, which further motivated probing the rainbow angle for context.
%
The cloud coverage fraction on ocean worlds could be constrained with observations spanning the ultraviolet to the near-infrared by finding the 
angle at which the planet's polarization phase curves at various wavelengths intersect \citep{treesandstam2019}.
%
While the glint feature can only be probed when a planet is at crescent phases, \citet{trees2022} showed that reflection by a liquid surface body can also be detected with spectropolarimetry when the planet is at phase angles where it is at the wide separations more easily probed when using coronagraphs.
%how use to map, 
%seen in Earthshine
%strong signal for exoplanets
% disentangled from cloud coupled with rainbow
%disentangled from lattitude albedo coupled in polarization

@@

\subsection{Rainbows and Glories}

Rainbows, glories and other optical effects from condensates provide a wealth of information about exoplanets \citep{karalidi2012rainbow, stam2008,2014A&A...566L...1G}.
%
In our own Solar System, the polarized back scatter from Titan provided evidence that the moon's reflection was due to haze rather than a surface \citep{zellner1973polarization}, on Venus the rainbow and phase curve behavior with wavelength showed the clouds were $\sim$75\% H$_2$SO$_4$ -- water mixture also constraining the droplet size distribution \citep{hansenhovenier1974}, and the glory was later used to rule out some solutions for the unknown UV absorber \citep{petrova2018glory}.

In unpolarized light the forward scattering and backscatter from clouds can dramatically change the shape of a phase curve from Lambertian.
%
While the glory and rainbow are technically visible in any phase curve, their strongly polarized nature coupled with the suppression of signal from other effects at gibbous and full phases in polarized light make them far easier to characterize in polarized light (both in the degree of polarization and in full Stokes measurements) \citep{karalidi2011, stam2008, treesandstam2019}.
%
The rainbow angle depends on the refractive index and therefore varies with wavelength and depending on the species.
%
The feature being far more narrow than the glint allows more exact determination of the species.
%
The droplet size distribution also effects the shape of the rainbow and clouds of species in a solid phase can diverge from the Mie scattering that produces these features greatly, producing other optical effects particularly at narrow scattering angles \citep{bailey2007, hansentravis1974, Karalidi2012ice}.
%
In this paper we define our optimal angle for rainbow detection as the peak of the water droplet rainbow in visible wavelengths.
%
However, extending observations to angles closer to the star (smaller scattering angles, i.e.\ larger phase angles) would allow the full shape of the rainbow to be characterized and ensure the characterization of rainbows from other possible HZ condensates.
%
The glory is produced when the droplet radii much smaller than the wavelength of light and can allow further characterization as their angle, intensity, and shape also vary with the complex refractive index but primarily occur within 20 degrees scattering angle very close to full phase. 
%
%The glory (in the backscattering direction) and diffraction peak (in the forward scattering direction) are distinct features in the total reflected flux, but not in the degree of polarization. In principle, the degree of polarization goes to zero in the forward and backward scattering directions. See e.g. Fig. 1 of \citet{stam2008}, Fig. 5 of \citet{trees2022}, Fig. 3 of \citet{karalidi2011} and \citet{hansentravis1974}.}}

@@

The National Academy of Sciences Astronomy \& Astrophysics 2020 Decadal Survey \citep{decadal} recommended the first in a new \enquote{Great Observatories} program a telescope with the capability to detect signatures of habitability on about 25 habitable zone planets.
%
This assertion requires an instrument with a coronagraph capable of high contrast imaging at optical to near infrared wavelengths.
%
Following this release, NASA recently announced the start of the development of the Habitable Worlds Observatory (\hwo).
%
The precursor technology recommended by the Decadal Survey also lists ``direct imaging to probe polarized ocean glint on terrestrial planets'' as one of the priority capabilities \citep[Box E.1 in][]{decadal} for ground and space-based observatories.
%
The performance and precise characteristics of this telescope (on-axis or off-axis, exact diameter, type of segmentation, type of coronagraph) are still to be determined.
%
The development will be heavily influenced by the LUVOIR \citep{LUVOIR2019} and HabEx \citep{HabEx_2020} preparatory studies.
%
The Inner Working angle (\IWA{}) of the resulting telescope will depend on these choices, which in turn will have a significant impact on the expected exo-Earth yield \citep{Stark2019_exoplanetyield}.

The Extreme Coronagraph for Living Planetary Systems (ECLIPS) coronograph for LUVOIR was a proposed \SIrange{200}{2000}{\nano\meter} instrument for exoplanet characterisation, with a similar coronagraph operating between \num{450} and \SI{1800}{\nano\meter} planned for HabEx.
%
POLLUX, a proposed UV instrument for the LUVOIR study could also detect  polarised light from hot Jupiters at \SI{300}{\nano\meter} \citep{Bouret2018_pollux}.
%
However, because of the challenges associated with the design of a coronagraph in the UV, it appears to be unlikely that \hwo{} will include a UV high-contrast instrument sensitive to Earth-like planets.

@@

In this paper, we quantify the planetary phase angles that would be accessible for direct observations for hypothetical terrestrial planets orbiting in the habitable zones of \hwo{} target stars.
%
In \S 2.1, we describe the target list, in \S 2.2 we derive expressions for the range of phase angles given a few different limits, in \S 2.3, we describe Monte Carlo simulations to marginalize over the unknown orbital inclination and eccentricity of planets in these systems, and in \S 2.4, we outline how the planet/star contrast changes as a function of the planet's orbital phase.  
%
We present and discuss our results in \S 3 and conclude in \S 4. It is important to note that the phases for each feature throughout this paper are approximate, that features overlap, and that the maximum peak can vary with composition and wavelength.
%the   the way that inclination and coronagraph inner working angle can limit the accessible phase angles  
% We minimally explore the contrast ratio as we are primarily exploring whether these scattering angles are reachable with inner working angle and outer working angle considered.

@@

\subsection{\hwo{} and Stellar Sample}
%\textcolor{blue}{Merge the 2 sections here into "Assumed technical specifications and sample"? I don't think this needs to be 2 subsections}
\hwo{} is still in the early stages of development, and the exact telescope and instrument design is as of yet undecided. 
%
In this work, we assume a \SI{6}{\meter} primary mirror, as suggested in the The National Academy of Sciences Astronomy \& Astrophysics 2020 Decadal Survey.
%
The observatory will utilise a coronagraph to suppress the stellar light and facilitate high-contrast observations of faint accompanying exoplanet. 
%
The \IWA{} of the coronagraph is currently undecided and will be driven by both the science requirements and technological limitations. 
%In this work, we investigate the scattering phase coverage obtainable for several choices for the \IWA{} of the observatory. 
%We also consider the Outer Working Angle (OWA) but for most targets this is not a limiting factor. 
%From the HabWorlds report. Inner Working Angle: The IWA defines the region near the star that cannot be accessed for direct imaging due to a coronagraphic mask, starshade obscuration, or an interferometric null. It depends on the architecture of the telescope, the starlight suppression system used, and the observation wavelength. While a nominal value could be defined by making assumptions about the design for Habitable Worlds Observatory, we choose instead to derive the IWA from the star list itself and Astro2020’s requirement that ~100 cumulative habitable zones be surveyed by direct imaging. As shown in the main table, an IWA near ~70 mas will be needed to access ~100 cumulative habitable zones, and would need to be achieved at all wavelengths of interest for spectral characterization. For any specific telescope and starlight suppression system architecture the IWA is typically proportional to wavelength; thus, the number of accessible HZs will strongly decrease as IWA(lambda) increases.
For ease of comparison, \cref{tab:IWA_OWA} lists the angular separations corresponding to multiples of $\lambda / D$ for a \SI{6}{\meter} telescope at representative visible and near infrared wavelengths of $\lambda = \SI{600}{\nano\meter}$ and $\SI{1}{\micro\meter}$.

\begin{table}
    \centering
    \caption{
        Conversion between $\lambda / D$ and \si{\mas} for different wavelengths, assuming $D = \SI{6}{\meter}$. 
        %\todo{We should use consistent wavelengths for this table, for the Bott Plot, and for the M\&M cumulative distribution function. Also update to use the only wavelength we choose.}
    }
    \label{tab:IWA_OWA}
    \begin{tabular}{ c c c c } 
    \toprule
     & $\lambda/D$ & mas (at \SI{600}{\nano\meter}) & mas (at \SI{1}{\micro\meter}) \\
    \midrule
    \midrule
    IWA & 1 & 20.63 & 34.38 \\
    IWA & 2 & 41.25 & 68.75 \\
    IWA & 3 & 61.88 & 103.13 \\
    IWA & 4 & 82.51 & 137.51 \\
    \midrule
    OWA & 32 &  660.05 & 1100.08 \\
    OWA & 64 & 1320.09 & 2200.16 \\
    \bottomrule
    \end{tabular}
\end{table}

\hwo{} is currently envisioned to observe the stars stated in the NASA Exoplanet Exploration Program's Mission Star List for the Habitable Worlds Observatory.
%
\footnote{\url{https://exoplanets.nasa.gov/internal_resources/2645_NASA_ExEP_Target_List_HWO_Documentation_2023.pdf}}
%
This list comprises $\sim$160 stars, the majority of which are Sun-like dwarfs; 66~F~dwarfs, 55~G~dwarfs, 40~K~dwarfs, and 3~M~dwarfs.
%
The target list is constructed assuming a maximum planet magnitude of $R_c = 31$ and a contrast floor of \num{2.5e-11}. 
%with an occurrence rate of rocky planets in the optimistic habitable zone to be $\eta_{\oplus}$ = 0.24 \citep{decadal}. 
The corresponding adopted habitable zone limits are a semi-major axis of \SIrange{0.95}{1.67}{\au} for a solar twin, planet sizes between \SIrange{0.8}{1.4}{} Earth radii, and for non-solar stars, scale as square root of the bolometric luminosity normalized to the Sun. 
%
This range of orbital separation corresponds to the \enquote{conservative habitable zone} \citep{kasting93, kopparapu13}. 
%
The authors of the report determined that these are the nearby stars (maximum distance \SI{25}{\parsec}) for which exo-Earths would be the most observable for an imaging survey of habitable zones with a \SI{6}{\meter} telescope.

In what follows, we consider hypothetical Earth-size planets orbiting at the right distance from each star to receive the same flux as the Earth receives from the Sun.
%
By construction, such planets should be observable at quadrature phase by HabWorlds.

@@

\subsection{Maximum scattering phase angle coverage}
\label{sec:Delta_phi}

The coronagraph can obscure part of the planet's orbit which, as demonstrated in \cref{fig:annotated-orbit}, can prevent the planet from being observed at high and low scattering angles. 
%
The range of scattering phase angles that can be observed will depend on the \IWA{} of the coronagraph as well as the shape and orientation of the exoplanet's orbit.

\begin{figure}
    \centering
    \includestandalone{tikz/annotated-orbit}
    \caption{
        An illustration of an edge-on circular orbit showing that the coronagraph can obscure the exoplanet at high and low scattering phase angles.
        %
        This can prevent observations of scattering phenomena which could be used to indicate the habitability and atmospheric composition of the exoplanet. The phase angle, $\varphi$, is shown in red and is the complement to the scattering angle in edge on systems.
        %\timmy{Maybe mention that this is a simplified illustration with a circular orbit, plus we are basically looking at an edge-on system?}
    }
    \label{fig:annotated-orbit}
\end{figure}

\subsubsection{Circular edge-on orbits}

>>>>

\usepackage{siunitx}
\sisetup{
    range-units = single,
    range-phrase = \textup{--},
}
\DeclareSIUnit\parsec{pc}
\DeclareSIUnit\au{au}
\DeclareSIUnit\mas{mas}

@@

% Define macros
\newcommand{\IWA}{\ensuremath{\mathrm{IWA}}}
\newcommand{\OWA}{\ensuremath{\mathrm{OWA}}}
\newcommand{\hwo}{HabWorlds}

@@

\begin{abstract}
NASA recently announced the Habitable Worlds Observatory (\hwo), a coronagraphic mission to detect rocky planets in habitable zones, establish their habitability, and search for biosignatures. 
%Numerous instrumental and observational challenges have still to be overcome to design a coronagraphic mission with that goal. After a research phase aiming at the detection of a few tens of exo-Earth in the habitable zone, the Habitable Worlds Observatory will use its spectroscopic and polarimetric capabilities to probe these planets' atmosphere and surface properties. 
Surface liquid water is central to the definition of planetary habitability.
%
Photometric and polarimetric phase variations can show the presence of specularly reflecting oceans on exoplanets, in addition to being sensitive to rainbows and other cloud scattering effects. 
%
Direct imaging missions are optimised to detect planets near quadrature, which may obscure some of the phases at which these features are strongest. 
%
The range of scattering phases accessible for an exoplanet can be limited by its orbital inclination or the coronagraph's inner working angle. 
%
Planets that orbit close to edge-on have accessible phase angles limited by obscuration by the coronagraph. 
%
We use the list of target stars for the Habitable Worlds Observatory to estimate the number of exo-Earths that could be searched for non-Lambertian scattering phenomena. 
%
We find that exo-Earths in systems listed in this Habitable Worlds Observatory catalog will have a detectable Rayleigh scattering peak. 
%
The glint signature at planetary phase angles of $\sim\qtyrange{50}{70}{\degree}$ 
%\sim$110--130$^\circ$ 
would be accessible to the Habitable Worlds Observatory in $\sim\num{14}$ systems, and the rainbow of water clouds at phases of \qtyrange{140}{160}{\degree} would be accessible in $\sim\num{42}$ systems, assuming a \qty{62}{\mas} inner working angle (\IWA{}; $3\,\lambda/D$ for a \qty{6}{\meter} telescope at \qty{600}{\nano\meter}).
%
Increasing the \IWA{} to $2\,\lambda/D$ increases the number of systems with detectable rainbows and glints by factors of roughly 2 and 3, respectively.
%
The number of systems in which an exo-Earth could be searched for these scattering phenomena increases with orbital eccentricity and scales approximately inversely with \IWA{}.  \end{abstract}

@@

%--------------------------------------------------------------------------------
\begin{figure*}
    \centering
    \includegraphics[width=\linewidth]{figures/Fig1IWAPol.png}
    \caption{
        An illustration of phase curves for two homogeneous planets in an edge on system with oceans and an Earth-like atmosphere. 
        The cloudy scenario includes an Earth-like water cloud deck at low altitudes. 
        Below the plot, we are showing the features to which the phase angles and scattering angles are the most sensitive. 
        In reality, as can be seen by the examples above, these features overlap.  
        Throughout the paper we approximate feature sensitivity ranges as those that capture the rise and fall of peaks in polarized light for most habitable zone species in the UV, Optical, and IR (scattering angles: glint $\sim\qtyrange{50}{70}{\degree}$; rainbow $\sim\SIrange{140}{160}{\degree}$; Rayleigh narrow of quadrature, and other forward and backscattering effects at the widest and narrowest scattering angles).
    }
    \label{fig:bottplot}
\end{figure*}
%--------------------------------------------------------------------------------

Signatures of habitability, from ocean glint to optical effects from condensates---like the rainbow---should be some of the most distinct signals in an exoplanet's reflected light phase curve, particularly in polarization.
%
However, depending on system architectures, some of these features can be obscured by the inner working angle of a coronagraph.

Liquid surface water is closely related to planetary habitability because it is essential for life as we know it.
%
To first order, the presence of liquid surface water can be predicted by the flux a planet receives from its star.
%
The habitable zone (HZ) is defined as the range of semi-major axes around a given main sequence star, for which a rocky planet (or moon) may have the right temperature to maintain liquid surface water \citep{kasting93}. 
%
However, there are many factors besides the stellar flux that influence a planet's surface temperature and ability to hold liquid surface water, such as the atmospheric surface pressure and, for example, geological activity. 
%
Establishing the presence of liquid surface water on an exoplanet would therefore provide important context for the interpretation of possible biosignatures.
%
Providing additional context to the habitability of an exoplanet is the presence of a water cycle, as evidenced by the identification of a rainbow and other optical effects that can be attributed to water clouds.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Exoplanet phase curves}

@@

\subsection{Ocean glint}

%As ocean glint is one of the few sources of specular reflection and is strongly polarized, it provides one of the best means to detect a signature of habitability in exoplanet phase curves. 
Ocean glint is the specular (Fresnel) reflection of light off smooth liquid surfaces.
%
In the total flux, the glint signal will be most prominent at small planetary phase (large scattering) angles.
%
The maximum polarization occurs at the Brewster angle determined by the refractive indices at the interface; between air and liquid water this occurs at \qty{53}{\degree} \citep[i.e., a scattering angle of \qty{127}{\degree}; see, e.g.,][]{2008Icar..195..927W}.
%
This angle therefore varies slightly with wavelength and with sea and atmosphere composition.
%
In general, the glint on a planet will be strongest at red wavelengths where the Rayleigh scattering in the planet's atmosphere and attenuation through the atmosphere are smallest \citep{Zugger_2011} and, conveniently, where the latitude-albedo effect (a degeneracy between polar brightness and glint) is also small \citep{2012ApJ...752L...3C}.
%
The strengthened glint signature at longer wavelengths has indeed been observed in Earthshine measurements \citep{Emde2017,sterzik2019, takahashi2021}.
%
{\color{red}{check the next sentence:}}
The broadening of the glint feature in the presence of wind-driven waves is well-described by the Cox-Munk solution \citep{CoxMunk1954}, shifts in both polarized and unpolarized light, and broadens across scattering angles in polarized light \citep{kopparla2018, Zugger_2010, treesandstam2019, trees2022}.

The glint provides evidence of liquid surface bodies, which, taken in the context of the temperature and types of condensates, may be interpreted as evidence for the presence of liquid water.
%
The forward scattering of light by cloud particles could be confused with glint \citep{Robinson_2010}, which further motivated probing the rainbow angle for context.
%
This confusion can be solved by measuring the intersection of phase curves at various wavelengths in polarized light (the polarized colour changes from blue, through white, to red), which only occurs when the glint contributes to the signal and at a planetary phase angle depending on the cloud coverage fraction \citep{treesandstam2019}.
%The cloud coverage fraction on ocean worlds could be constrained with observations spanning the ultraviolet to the near-infrared by finding the 
%angle at which the planet's polarization phase curves at various wavelengths intersect \citep{treesandstam2019}.
%
While the glint feature can only be probed when a planet is at crescent phases, \citet{trees2022} showed that reflection by a liquid surface body can also be detected with spectropolarimetry when the planet is at phase angles where it is at the wide separations more easily probed when using coronagraphs.
%how use to map, 
%seen in Earthshine
%strong signal for exoplanets
% disentangled from cloud coupled with rainbow
%disentangled from lattitude albedo coupled in polarization

@@

\subsection{Rainbows and glories}

Rainbows, glories and other optical effects from condensates provide a wealth of information about exoplanets \citep{karalidi2012rainbow, stam2008,2014A&A...566L...1G}.
%
In our own Solar System, the polarized back scatter from Titan provided evidence that the moon's reflection was due to haze rather than a surface \citep{zellner1973polarization}, the phase curve behaviour of Venus showed not only that the clouds were $\sim$75\% H$_2$SO$_4$--water mixture but also constrained the droplet size distribution \citep{hansenhovenier1974}. The glory feature in Venus's phase curve was later used to rule out some solutions for the unknown UV absorber \citep{petrova2018glory}.

In unpolarized light, the scattering by cloud particles can dramatically change the shape of a planet's phase curve compared to that of a 
Lambertian reflecting planet.
%
{\color{red}{check the sentence:}}
While the glory and rainbow are technically visible in any phase curve
in the presence of spherical cloud and/or haze particles, their strongly polarized nature coupled with the suppression of signals from other effects at gibbous and full phases in polarized light make them far easier to characterize in polarized light (both in the degree of polarization and in full Stokes measurements) \citep{karalidi2011, stam2008, treesandstam2019}.
%
The rainbow angle depends on the refractive index of the scatterers and thus depends on the composition and the wavelength. 
The wavelength dependence itself also depends on the particles' size distribution, and the size distribution affects the angular shape of the rainbow \citep{karalidi2011}
%
Because the rainbow feature is more narrow than the glint, it would 
allow for a more exact determination of the particle composition.
%
Note that the scattering by clouds that consist of solid particles, 
such as ice crystals, will produce other optical features, such as halo's, particularly at small scattering angles \citep{bailey2007, hansentravis1974, Karalidi2012ice}.
%

In this paper, we define our optimal angle for the detection of a rainbow as that of the peak of the water droplet rainbow at visible wavelengths.
%
Pushing observations to angles closer to the star (smaller scattering angles, i.e., larger phase angles) would allow the determination of the full shape of this rainbow, and enable the identification of rainbows due to the scattering of light by other possible HZ condensates.
%
The glory 
%is produced when the droplet radii much smaller than the wavelength of light 
can allow further characterization of condensates as its angle and shape also depend on the complex refractive index, but this feature occurs within \qty{20}{\degree} scattering angle very close to full phase \citep{hansentravis1974}.
%
%The glory (in the backscattering direction) and diffraction peak (in the forward scattering direction) are distinct features in the total reflected flux, but not in the degree of polarization. In principle, the degree of polarization goes to zero in the forward and backward scattering directions. See e.g. Fig. 1 of \citet{stam2008}, Fig. 5 of \citet{trees2022}, Fig. 3 of \citet{karalidi2011} and \citet{hansentravis1974}.}}

@@

The National Academy of Sciences Astronomy \& Astrophysics 2020 Decadal Survey \citep{decadal} recommended the first in a new \enquote{Great Observatories} program, a telescope with the capability to detect signatures of habitability on about 25 HZ planets.
%
This requires an instrument with a coronagraph capable of high contrast imaging at optical to near infrared wavelengths.
%
Following the release of this survey, NASA recently announced the start of the development of the Habitable Worlds Observatory (\hwo).
%

The precursor technology recommended by the survey lists \enquote{direct imaging to probe polarized ocean glint on terrestrial planets} as one of the priority capabilities \citep[Box E.1 in][]{decadal} for ground and space-based observatories.
%
The performance and precise characteristics of \hwo{} (on-axis or off-axis, diameter, type of segmentation, type of coronagraph) are still to be determined.
%
The development will be heavily informed by the LUVOIR \citep{LUVOIR2019} and HabEx \citep{HabEx_2020} preparatory studies.
%
The Inner Working angle (\IWA{}) of the resulting telescope will depend on these characteristics, which in turn will have a significant impact on the expected exo-Earth yield \citep{Stark2019_exoplanetyield}.

The Extreme Coronagraph for Living Planetary Systems (ECLIPS) coronograph for LUVOIR was a proposed \qtyrange{200}{2000}{\nano\meter} instrument for exoplanet characterization, with a similar coronagraph, albeit operating between \num{450} and \qty{1800}{\nano\meter}, planned for HabEx.
%
POLLUX, a proposed instrument for LUVOIR would detect polarized light from hot Jupiters at \qty{300}{\nano\meter} \citep{Bouret2018_pollux}.
%
However, because of the challenges associated with the design of a coronagraph in the UV, it appears to be unlikely that \hwo{} will include a UV high-contrast instrument sensitive to Earth-like planets.

@@

In this paper, we quantify the planetary phase angles that would be accessible for direct observations for hypothetical terrestrial planets orbiting in the habitable zones of \hwo{} target stars.
%
In \cref{subsec:2.1}, we describe the target list, in \cref{subsec:2.2} we derive expressions for the range of phase angles given a few different limits, in \cref{subsec:2.3}, we describe Monte Carlo simulations to marginalize over the unknown orbital inclination and eccentricity of planets in these systems, and in \cref{subsec:2.4}, we outline how the planet/star contrast changes as a function of the planet's orbital phase.  
%
We present and discuss our results in \cref{sec:3}, and conclude in \cref{sec:4}. 
It is important to note that the phases for the angular features throughout this paper are approximate, that features overlap, and that their maximum peak will usually vary with particle composition and wavelength.
%the   the way that inclination and coronagraph inner working angle can limit the accessible phase angles  
% We minimally explore the contrast ratio as we are primarily exploring whether these scattering angles are reachable with inner working angle and outer working angle considered.

@@

\subsection{\hwo{} and its stellar sample}
\label{subsec:2.1}

%\textcolor{blue}{Merge the 2 sections here into "Assumed technical specifications and sample"? I don't think this needs to be 2 subsections}
\hwo{} is still in the early stages of development, and although the exact telescope and instrument design have not yet been determined, we assume a \qty{6}{\meter} primary mirror, as suggested in The National Academy of Sciences Astronomy \& Astrophysics 2020 Decadal Survey.
%
The observatory will utilise a coronagraph to suppress the stellar light and facilitate high-contrast observations of faint accompanying exoplanets. 
%
The \IWA{} of the coronagraph will be driven by both the science requirements and technological limitations and is still unknown.
%In this work, we investigate the scattering phase coverage obtainable for several choices for the \IWA{} of the observatory. 
%We also consider the Outer Working Angle (OWA) but for most targets this is not a limiting factor. 
%From the HabWorlds report. Inner Working Angle: The IWA defines the region near the star that cannot be accessed for direct imaging due to a coronagraphic mask, starshade obscuration, or an interferometric null. It depends on the architecture of the telescope, the starlight suppression system used, and the observation wavelength. While a nominal value could be defined by making assumptions about the design for Habitable Worlds Observatory, we choose instead to derive the IWA from the star list itself and Astro2020’s requirement that ~100 cumulative habitable zones be surveyed by direct imaging. As shown in the main table, an IWA near ~70 mas will be needed to access ~100 cumulative habitable zones, and would need to be achieved at all wavelengths of interest for spectral characterization. For any specific telescope and starlight suppression system architecture the IWA is typically proportional to wavelength; thus, the number of accessible HZs will strongly decrease as IWA(lambda) increases.
For ease of comparison, \cref{tab:IWA_OWA} lists the angular separations corresponding to multiples of $\lambda / D$ for a \qty{6}{\meter} telescope at representative wavelengths of $\lambda = \qty{600}{\nano\meter}$ and $\qty{1}{\micro\meter}$.

%--------------------------------------------------------------------------
\begin{table}
    \centering
    \caption{
        Conversion between $\lambda / D$ and \unit{\mas} for wavelengths $\lambda$ of \qty{600}{\nano\meter} and \qty{1}{\micro\meter}, assuming the telescope's diameter $D$ is $\qty{6}{\meter}$ for the Inner and Outer Working Angles (i.e., \IWA{} and \OWA{}, respectively). 
        %\todo{We should use consistent wavelengths for this table, for the Bott Plot, and for the M\&M cumulative distribution function. Also update to use the only wavelength we choose.}
    }
    \label{tab:IWA_OWA}
    \begin{tabular}{ c c c c } 
    \toprule
    & $\lambda/D$ & \unit{\mas} (at \qty{600}{\nano\meter}) & \unit{\mas} (at \qty{1}{\micro\meter}) \\
    \midrule
    \midrule
    IWA & 1 & 20.63 & 34.38 \\
    IWA & 2 & 41.25 & 68.75 \\
    IWA & 3 & 61.88 & 103.13 \\
    IWA & 4 & 82.51 & 137.51 \\
    \midrule
    OWA & 32 &  660.05 & 1100.08 \\
    OWA & 64 & 1320.09 & 2200.16 \\
    \bottomrule
    \end{tabular}
\end{table}
%--------------------------------------------------------------------------

\hwo{} is envisioned to observe the stars in the NASA Exoplanet Exploration Program's Mission Star List for the Habitable Worlds Observatory.%
\footnote{\hwo{} target list: \url{https://exoplanets.nasa.gov/internal_resources/2645_NASA_ExEP_Target_List_HWO_Documentation_2023.pdf}}
%
This list comprises $\sim\num{160}$ stars, the majority of which are Sun-like dwarfs: 66~F~dwarfs, 55~G~dwarfs, 40~K~dwarfs, and 3~M~dwarfs.
%
The target list is constructed assuming a maximum planet magnitude of $R_c = 31$ and a contrast floor of \num{2.5e-11}. 
%with an occurrence rate of rocky planets in the optimistic habitable zone to be $\eta_{\oplus}$ = 0.24 \citep{decadal}. 
The corresponding adopted habitable zone limits are a semi-major axis of \qtyrange{0.95}{1.67}{\au} for a solar twin, planet sizes between \qtyrange{0.8}{1.4}{} Earth radii, and for non-solar stars, scale as square root of the bolometric luminosity normalized to the Sun. 
%
This range of orbital separation corresponds to the \enquote{conservative habitable zone} \citep{kasting93, kopparapu13}. 
%
The authors of the Decadal Survey determined that for an imaging survey of habitable zones with a \qty{6}{\meter} telescope, these are the nearby stars (maximum distance \qty{25}{\parsec}) around which exo-Earths would be the best observable.

In what follows, we consider hypothetical Earth-sized planets orbiting at the right distance from their parent star to receive the same flux as the Earth receives from the Sun.
%
By design, such planets should be observable by \hwo{} at least at their quadrature phase.

@@

\subsection{Maximum scattering phase angle coverage}
\label{subsec:2.2}

A coronagraph can obscure part of the planet's orbit which, as shown in \cref{fig:annotated-orbit}, can prevent the planet from being observable at high and low scattering angles. 
%
The range of observable scattering phase angles will depend on the \IWA{} of the coronagraph, as well as on the orientation and shape of the exoplanet's orbit.

%--------------------------------------------------------------------------
% Figure 2
%--------------------------------------------------------------------------
\begin{figure}
    \centering
    \includestandalone{tikz/annotated-orbit}
    \caption{
        An illustration of a planet in an edge-on circular orbit showing how a coronagraph can obscure the exoplanet at high and low scattering phase angles, and thus prevent observations of scattering features in an exoplanet's phase curve that could help to identify its atmospheric and/or surface composition and habitability. 
        The phase angle, $\varphi$, is shown in red and is the complement of the scattering angle in edge on systems.
    }
    \label{fig:annotated-orbit}
\end{figure}
%--------------------------------------------------------------------------

\subsubsection{Circular, edge-on orbits}
